---
title: "Bike Sharing"
author: "Ramaa Nathan"
date: "7/25/2018"
#output: html_document
always_allow_html: yes
output: 
  github_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(utils) #to read csv files
library(ggplot2) #for ggplot
library(base) #for strptime
library(car) #for vif
library(MASS)
library(caret) #for cross validations
library(knitr)
library(gam) #for gam models
library(randomForest)
library(gbm) #for boosting
library(plyr) #for do.call 
```
```
Project Goal
You are provided hourly rental data spanning two years. For this competition, the training set (train.csv) is comprised of the first 19 days of each month, while the test set (test.csv) is the 20th to the end of the month. You must predict the total count of bikes rented during each hour covered by the test set, using only information available prior to the rental period.
```

## Exploratory Data Analysis

```{r}
bikeTrain = read.csv("train.csv",header=TRUE,sep=",")
str(bikeTrain)
bikeTest = read.csv("test.csv",header=TRUE,sep=",")
str(bikeTest)

```

```
Data Descriptions:
   1. datetime - hourly date + timestamp  
   2. season - (actually categorical - ordinal) 
      1 = spring, 2 = summer, 3 = fall, 4 = winter 
   3. holiday - whether the day is considered a holiday
   4. workingday - whether the day is neither a weekend nor holiday
   5. weather - (categorical - ordinal)
      1: Clear, Few clouds, Partly cloudy, Partly cloudy 
      2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist 
      3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds 
      4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog 
   6. temp - temperature in Celsius
   7. atemp - "feels like" temperature in Celsius (function of humidity and windspeed)
   8. humidity - relative humidity
   9. windspeed - wind speed
   10. casual - number of non-registered user rentals initiated
   11. registered - number of registered user rentals initiated
   12. count - number of total rentals

Casual, Registered, and Count are all response variables. Count is sum of Casual+Registered

The  pair of indicator variables, holiday and working day, together indicate if the day is a working day, holiday, or a weekend.

```
```{r}
#For data exploration purposes, we will combine the training and test data sets.
bikeRentals = as.data.frame(rbind(bikeTrain[,1:9],bikeTest[]))
str(bikeRentals)
```

Check for missing data
```{r}
summary(bikeRentals)
#check for na
any(is.na(bikeRentals))
#check for duplicates
any(duplicated(bikeRentals))
```


```{r}
#check distribution of all the numeric data in the combined data sets
par(mfrow=c(3,3))
for (i in which(sapply(bikeRentals,is.numeric))) {
  hist(bikeRentals[,i],freq=TRUE,xlab=colnames(bikeRentals)[i],main="")
}
par(mfrow=c(1,1))
```

```{r}
#check for combinations of holiday and workginday
with(bikeRentals,unique(cbind(holiday,workingday)))
#number of entries for holiday (note this includes multiple entries per day)
with(bikeRentals,sum(holiday))
#number of entries for holiday (note this includes multiple entries per day)
with(bikeRentals,sum(workingday))
```

Season is categorical (nominal) varaible indicate the four seasons.
Holiday - Binary variable - is it a public holiday? - Does not include weekends
Working Day - one of the five working days in a week, which are not holidays. So, a day that is neither a holiday nor a working day can be considered to be a weekend
```{r}
#separate out the time from timestamp
#extract the hour in each observation
bikeTrain$datetime <- strptime(bikeTrain$datetime, format="%Y-%m-%d %H:%M:%S")
bikeTrain$hour = bikeTrain$datetime$hour

#put hours into bins
bikeTrain$hourBins = cut(bikeTrain$hour,c(0,6,10,13,16,20,24),labels=c("0-5","6-9","10-12","13-15","16-19","20-24"),right=FALSE)
```


###Plots
```{r}
ggplot(bikeTrain,aes(x=season,y=count)) + 
  geom_point() + 
  ggtitle("Rental vs Season") +  
  scale_x_discrete(limits=c("Spring","Summer","Fall","Winter"))+
  labs(x="Season", y="Count of bikes rented")
```
```
observation - bikes are rented every season - though there seem to be a few days in the fall when more bikes are rented.
```
```{r}
ggplot(bikeTrain,aes(x=weather,y=count)) + 
  geom_point() + 
  ggtitle("Rental vs Weather") +  
  scale_x_discrete(limits=c("Clear","Cloudy","Light Rain/Snow", "Snow/Heavy Rain")) +
  labs(x="Weather", y="Count of bikes rented")
```
```
observation - bike rental very much depends on weather. Highest bike rentals occur during clear weather and almost no bikes are rented during really bad or inclement weather
```
```{r}
# - 1:weekend; 2:holiday; 3:workingday
ggplot(bikeTrain,aes(x=holiday+(2*workingday)+1,y=count)) + 
  geom_point() + 
  ggtitle("Rental vs Type of Day") +  
  scale_x_discrete(limits=c("Weekend","Holiday","Working Day")) +
  labs(x="Type of Day", y="Count of bikes rented")
  
```
```
observation:  many more bikes are rented on working days than on a weekend. Least number of bikes are rented on holidays.
```

```{r}
ggplot(bikeTrain,aes(x=temp,y=atemp)) + 
  geom_point() + 
  ggtitle("Actual Temp vs Feel Temp") +
  labs(x="Actual Temperature", y="Feeling Temperature") 

ggplot(bikeTrain,aes(x=atemp,y=count,colour=bikeTrain$hourBins)) + 
  geom_point() + 
  ggtitle("Count vs Feel Temp") +
  labs(x="Actual Temperature", y="Count") 
```
```
observation: When the temperature is between 15degC and 25 degC, there is no difference between the actual and feeling temperature. There is a variation between the two types of temperature at higher and lower temperatures. But, it also evident that the two temperatures are highly correlated. So, while modeling, we should consider only one of them. As the bike sharing is an outdoor activity, it may make sense to use the atemp or feeling temperature.
```
```{r}
ggplot(bikeTrain,aes(x=casual,y=registered)) + 
  geom_point() + 
  ggtitle("Casual vs Registered Renters") +
  labs(x="Unregistered Renters", y="Registered Renters") 
```

```{r}
#bikeTrain$Hour = substr(bikeTrain$datetime, 12, 13) # extract Hour of rental
bikeTrain$TypeOfDay = bikeTrain$holiday+(2*bikeTrain$workingday)+1
bikeTrain$TypeOfDay = as.factor(bikeTrain$TypeOfDay)
levels(bikeTrain$TypeOfDay)=c("Weekend","Holiday","Working Day")
ggplot(bikeTrain,aes(x=hour,y=count,colour=TypeOfDay)) + 
  geom_point() + 
  ggtitle("Rental by Time of Day") +
  labs(x="Hour of Day", y="Count of bikes rented") 
```
```
Observation: When observed by the hour of the day, bikes are rented at all hours of the day with the highest number of bikes being rented at the 8am hour and from 5pm to 7pm on working days. This clearly indicates that bikes are rented as commutes to and from work.
```
```{r}
weatherF = as.factor(bikeTrain$weather)
levels(weatherF)=c("Clear","Cloudy","Light Rain/Snow", "Snow/Heavy Rain")
ggplot(bikeTrain,aes(x=hour,y=count,colour=weatherF)) + 
  geom_point() + 
  ggtitle("Rental by Time of Day") +
  labs(x="Hour of Day", y="Count of bikes rented") 
```
```
Observation: When observed by the hour of the day, bikes are rented during late hours of the day when the weather is either clear or cloudy. Bikes are rented during light rain only during the middle of the day.
```
```{r}
seasonF = as.factor(bikeTrain$season)
levels(seasonF)=c("Spring","Summer","Fall","Winter")
ggplot(bikeTrain,aes(x=hour,y=count,colour=seasonF)) + 
  geom_point() + 
  ggtitle("Rental by Time of Day") +
  labs(x="Hour of Day", y="Count of bikes rented") 
```
```
Observation - It seems counter-intuitive that mode number of bikes are rented during the winter than in spring
```
```
Based on all the observations, there seems to be an interaction between the hours of the day, the weather, and whether it is a holiday. This will need to be verified using the different models

Approaches:
   1. Multiple Linear Regression
      a. Do Residual Analysis
   2. Feature Reductin - using subseet regression
   3. Apply Regularization 
      a. Ridge
      b. Lasso
   4. PCR
   5. PLS
   6. Random Forests
   7. Boosting
   8. Additive Models
   
In all cases, apply cross validation and pick the best values.
Compare all the errors and determine the accuracy

*B IMportant to note that we can predict only using the previous training values *B

```
```{r}
#extract the dates -- playing around with the strptime function - this subsection not needed for analysis
dates = substr(bikeTrain$datetime, 1, 10) # extract Hour of rental
uniqueDates = unique(dates) 
#uniqueDates

bikeTrain$datetime <- strptime(bikeTrain$datetime, format="%Y-%m-%d %H:%M:%S")

bikeTrain$year = bikeTrain$datetime$year
bikeTrain$mon = bikeTrain$datetime$mon #Jan is represented as 0, Dec as 11 - months() - prints name
bikeTrain$mday = bikeTrain$datetime$mday #0-31 -- 
bikeTrain$hour = bikeTrain$datetime$hour #0-23 hours 
#head(bikeTrain$datetime$min) #0-59  minutes
#head(bikeTrain$datetime$sec) #0-61 seconds
bikeTrain$wday = bikeTrain$datetime$wday #0-6 das of the week: weekdays() prints name
```

```{r}

```


```{r}
#Check correlations between the numeric variables.
 cor(cbind(temp=bikeTrain$temp,atemp=bikeTrain$atemp,humidity=bikeTrain$humidity,windspeed=bikeTrain$windspeed))

```
```
Observation: Temp and atemp are highy correlated - causing multicollinearity. So, we would not want to include both variables in the model. As perople tend to do outdoor activities by the "feeling temperature" rather the actual, we will only use atemp.
```

## Data Modeling



```{r 1}
bikeTrain = read.csv("train.csv",header=TRUE,sep=",")
bikeTest = read.csv("test.csv",header=TRUE,sep=",")
str(bikeTrain)
```

Extract the details of the time variables
```{r 2}
bikeTrain$datetime <- strptime(bikeTrain$datetime, format="%Y-%m-%d %H:%M:%S")
bikeTrain$year = bikeTrain$datetime$year
bikeTrain$mon = bikeTrain$datetime$mon #Jan is represented as 0, Dec as 11 - months() - prints name
#create factor variables of year and month
bikeTrain$yearf=as.factor(bikeTrain$year)
bikeTrain$monf=as.factor(bikeTrain$mon)
bikeTrain$hour = as.factor(bikeTrain$datetime$hour)
bikeTrain$mday = as.factor(bikeTrain$datetime$mday) #0-31 -- 
#bikeTrain$mday = bikeTrain$datetime$mday
bikeTrain$hour = as.factor(bikeTrain$datetime$hour) #0-23 hours 
#head(bikeTrain$datetime$min) #0-59  minutes
#head(bikeTrain$datetime$sec) #0-61 seconds
bikeTrain$wday = as.factor(bikeTrain$datetime$wday) #0-6 das of the week: weekdays() prints name
```


Convert categorical predictors from numerical to factor data type
```{r 3}
#additional columns
#create factor variables for train set
bikeTrain$typeofday = as.factor(bikeTrain$holiday+(2*bikeTrain$workingday)+1)
levels(bikeTrain$typeofday)=c("Weekend","Holiday","Working Day")
bikeTrain$season = factor(bikeTrain$season)
levels(bikeTrain$season)=c("Spring","Summer","Fall","Winter")
bikeTrain$weather = as.factor(bikeTrain$weather)
levels(bikeTrain$weather)=c("Clear","Cloudy","Light Rain/Snow", "Snow/Heavy Rain")
#str(bikeTrain)
#remove unwanted columns
bikeTrain = within(bikeTrain,rm(holiday,workingday,casual,registered,datetime,temp))
bikeTrain$name = ""
str(bikeTrain)
#knit_exit()
```

Create and transform the test dataset now
```{r 4}
#remove entry with first in the year 2011
bikeTest=bikeTest[-which(row.names(bikeTest)==586),]

bikeTest$datetime <- strptime(bikeTest$datetime, format="%Y-%m-%d %H:%M:%S")
bikeTest$year = bikeTest$datetime$year
bikeTest$mon = bikeTest$datetime$mon #Jan is represented as 0, Dec as 11 - months() - prints name
#create factor variables of year and month
bikeTest$yearf=as.factor(bikeTest$year)
bikeTest$monf=as.factor(bikeTest$mon)
bikeTest$mday = as.factor(bikeTest$datetime$mday) #0-31 -- 
#bikeTest$mday = bikeTest$datetime$mday #0-31 -- 
bikeTest$hour = as.factor(bikeTest$datetime$hour) #0-23 hours 
#head(bikeTest$datetime$min) #0-59  minutes
#head(bikeTest$datetime$sec) #0-61 seconds
bikeTest$wday = as.factor(bikeTest$datetime$wday) #0-6 das of the week: weekdays() prints name

#Convert categorical predictors from numerical to factor data type
#additional columns
#create factor variables for Test set
bikeTest$typeofday = as.factor(bikeTest$holiday+(2*bikeTest$workingday)+1)
levels(bikeTest$typeofday)=c("Weekend","Holiday","Working Day")
bikeTest$season = factor(bikeTest$season)
levels(bikeTest$season)=c("Spring","Summer","Fall","Winter")
bikeTest$weather = as.factor(bikeTest$weather)
levels(bikeTest$weather)=c("Clear","Cloudy","Light Rain/Snow", "Snow/Heavy Rain")
#remove unwanted columns
bikeTest = within(bikeTest,rm(holiday,workingday,casual,registered,datetime,temp))
bikeTest$name = ""
bikeTest$count = rep(0,nrow(bikeTest))
#str(bikeTest)
```



```{r 5}
#read the data from the UCI dataset that contains the counts for the testset too
fulldata = read.csv("hour_fullset.csv",header=TRUE,sep=",",stringsAsFactors=FALSE)
bikeTest2 = fulldata[which(strptime(fulldata$datetime, format="%Y-%m-%d %H:%M:%S")$mday > 19),]
#remove entry with snow in the first month of 2011
bikeTest2=bikeTest2[-which(row.names(bikeTest2)==586),]


#Create and transform the new test dataset now
bikeTest2$datetime <- strptime(bikeTest2$datetime, format="%Y-%m-%d %H:%M:%S")
bikeTest2$year = bikeTest2$datetime$year
bikeTest2$mon = bikeTest2$datetime$mon #Jan is represented as 0, Dec as 11 - months() - prints name
#create factor variables of year and month
bikeTest2$yearf=as.factor(bikeTest2$year)
bikeTest2$monf=as.factor(bikeTest2$mon)
bikeTest2$hour = as.factor(bikeTest2$datetime$hour)
bikeTest2$mday = as.factor(bikeTest2$datetime$mday) #0-31 -- 
#bikeTest2$mday = bikeTest2$datetime$mday #0-31 -- 
bikeTest2$hour = as.factor(bikeTest2$datetime$hour) #0-23 hours 
#head(bikeTest2$datetime$min) #0-59  minutes
#head(bikeTest2$datetime$sec) #0-61 seconds
bikeTest2$wday = as.factor(bikeTest2$datetime$wday) #0-6 das of the week: weekdays() prints name

#Convert categorical predictors from numerical to factor data type
#additional columns
#create factor variables for Test set
bikeTest2$typeofday = as.factor(bikeTest2$holiday+(2*bikeTest2$workingday)+1)
levels(bikeTest2$typeofday)=c("Weekend","Holiday","Working Day")
bikeTest2$season = factor(bikeTest2$season)
levels(bikeTest2$season)=c("Spring","Summer","Fall","Winter")
bikeTest2$weather = as.factor(bikeTest2$weather)
levels(bikeTest2$weather)=c("Clear","Cloudy","Light Rain/Snow", "Snow/Heavy Rain")
#remove unwanted columns
bikeTest2 = within(bikeTest2,rm(holiday,workingday,casual,registered,datetime,temp))
#remove row with snow in jan 2011


str(bikeTest2)
#knit_exit()
```

```{r 6}
#create the multiple datasets
unique(bikeTrain$yearf)
unique(bikeTrain$monf)
#split the train data by the factor variables yearf, monf
bikeTrainSets = split(bikeTrain,list(bikeTrain$yearf,bikeTrain$monf))

#create the strings on which the split was done to help build the growing sets of datasets
#combine year and mon - cannot use factor variables directly in cbind
yearmon = cbind(bikeTrain$year,bikeTrain$mon)
#yearmon = cbind(levels(bikeTrain$yearf),levels(bikeTrain$monf))
#order the columsn first by year and then by column
yearmons = unique(yearmon[order(yearmon[,1],yearmon[,2],decreasing = FALSE),])
#yearmons
#combine the ordered entries into a string
splitString=paste(yearmons[,1],'.',yearmons[,2],sep="")
#splitString

#create an incremental list of dataframes - incremental in the data elements
bikeTrainIncrList = list() 
prevBikeTrain = data.frame()
for(i in splitString) {
  #print(bikeTrainSets[[i]])
  bikeTrainIncrList[[i]] = rbind(prevBikeTrain,bikeTrainSets[[i]])
  prevBikeTrain = bikeTrainIncrList[[i]]
  bikeTrainIncrList[[i]]$name = i
}

#check that each element in the list is a dataframe and combines the data from the previous months
# start of print functions
#sapply(bikeTrainIncrList,FUN=str)
#sapply(bikeTrainIncrList,FUN=function(x) unique(cbind(x$yearf,x$monf)))
#length(bikeTrainIncrList)
#end of print functions
#knit_exit()
```


```{r 7}
#for each entry in biketest, determine the corresponding model that needs to be used
getYearMon = function(x) {
  #print(paste(trimws(x["year"]),'.',trimws(x["mon"]),sep=""))
  return(paste(trimws(x["year"]),'.',trimws(x["mon"]),sep=""))
}
set.seed(1)
testSample = bikeTest[sample(nrow(bikeTest),30),]
#apply(testSample, MARGIN=1, FUN=function(x) getYearMon(x))
matchTrainList = apply(testSample, MARGIN=1, FUN=function(x) bikeTrainIncrList[[getYearMon(x)]])

#start of print functions....
#str(matchTrainList)
#sapply(seq(30),FUN=function(x) cbind(testSample$year[x],testSample$mon[x],unique(cbind(matchTrainList[[x]]$year,matchTrainList[[x]]$mon))))#
#end of print functions...

#check the same for bikeTest2
testSample2 = bikeTest2[sample(nrow(bikeTest2),30),]
matchTrainList2 = apply(testSample2, MARGIN=1, FUN=function(x) bikeTrainIncrList[[getYearMon(x)]])

#start of print functions....
#str(matchTrainList)
#sapply(seq(30),FUN=function(x) cbind(testSample2$year[x],testSample2$mon[x],unique(cbind(matchTrainList2[[x]]$year,matchTrainList2[[x]]$mon))))#
#end of print functions ....
#knit_exit()
```


```{r 8}
#apply data models  - lm
 func_fit_lm = function(trainset,testset,factorpreds,dname) {
   #create the formula string
   formula_string = paste("count~",factorpreds,"+atemp+humidity+windspeed",sep="")
   print(formula_string)
  fit_lm = lm(as.formula(formula_string),data=trainset)
  trainerror = mean((trainset$count - predict(fit_lm,newdata=trainset))^2)
  testerror = mean((testset$count - predict(fit_lm,newdata=testset))^2)
  #print(c(trainerror,testerror))
  cat("dataset=",dname,"model=lm","trainerror=",trainerror,"testerror=",testerror)
  
  #analyse the result
  summary(fit_lm)
  #vif(fit_lm)
  anova(fit_lm)
  par(mfrow=c(2,2))
  plot(fit_lm)
  par(mfrow=c(1,1))
  #print(list("model"="lm","trainerror"=trainerror,"testerror"=testerror))
  return(list("dataset"=unique(trainset$name),"model"="lm","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_lm))
    
 }
```


```{r 9}
 func_fit_gam_lm = function(trainset,testset,factorpreds,dname) {
   #print(dim(trainset))
   #create the formula string
   formula_string = paste("count~",factorpreds,"+atemp+humidity+windspeed",sep="")
   print(formula_string)
   
   fit_gam_lm = gam(as.formula(formula_string),data=trainset)
   print(summary(fit_gam_lm))
   
   trainerror = mean((trainset$count - predict(fit_gam_lm,newdata=trainset)) ^2)
   testerror = mean((testset$count - predict(fit_gam_lm,newdata=testset))^2)
 
   cat("dataset=",dname,"model=gam_lm","trainerror=",trainerror,"testerror=",testerror)
   plot.Gam(fit_gam_lm, se=TRUE, col="green")

   return(list("dataset"=unique(trainset$name),"model"="gam_lm","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_gam_lm))
 }
```


```{r 10}
 func_fit_gam_poly = function(trainset,testset,factorpreds,dname) {
   print(dim(trainset))
   formula_string_2 = paste("count~",factorpreds,"+poly(atemp,2)+poly(humidity,2)+poly(windspeed,2)",sep="")
   fit_gam_poly_2 = gam(as.formula(formula_string_2),data=trainset)
   
   formula_string_3 = paste("count~",factorpreds,"+poly(atemp,3)+poly(humidity,3)+poly(windspeed,3)",sep="")
   fit_gam_poly_3 = gam(as.formula(formula_string_3),data=trainset)
  
    formula_string_4 = paste("count~",factorpreds,"+poly(atemp,4)+poly(humidity,4)+poly(windspeed,4)",sep="")
   fit_gam_poly_4 = gam(as.formula(formula_string_4),data=trainset)
   
   formula_string_5 = paste("count~",factorpreds,"+poly(atemp,5)+poly(humidity,5)+poly(windspeed,5)",sep="")
   fit_gam_poly_5 = gam(as.formula(formula_string_5),data=trainset)
   
   #compare the different poly models
   print(summary(fit_gam_poly_2))
   print(summary(fit_gam_poly_3))
   print(summary(fit_gam_poly_4))
   print(summary(fit_gam_poly_5))
   print(anova(fit_gam_poly_2,fit_gam_poly_3,fit_gam_poly_4,fit_gam_poly_5))
   #observation - ploy_3 seems to be the best fit in general

   trainerror = mean((trainset$count - predict(fit_gam_poly_3,newdata=trainset))^2)
   testerror = mean((testset$count - predict(fit_gam_poly_3,newdata=testset))^2)

  cat("dataset=",dname,"model=gam_poly","trainerror=",trainerror,"testerror=",testerror)

   return(list("dataset"=unique(trainset$name),"model"="gam_poly","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_gam_poly_3))
}
```


```{r 11}
 func_fit_gam_s = function(trainset,testset,factorpreds,dname) {
    formula_string = paste("count~",factorpreds,"+s(atemp)+s(humidity)+s(windspeed)",sep="")
  print(formula_string)
    fit_gam_s = gam(as.formula(formula_string),data=trainset)
   trainerror = mean((trainset$count-predict(fit_gam_s,newdata=trainset))^2)
   testerror = mean((testset$count-predict(fit_gam_s,newdata=testset))^2)
   #print(head(cbind(trainset$mday,trainset$hour,trainset$count,predict(fit_gam_s,newdata=trainset))),20)
   #print(head(cbind(testset$mday,testset$hour,testset$count,predict(fit_gam_s,newdata=testset))),20)
   
   cat("dataset=",dname,"model=gam_s","trainerror=",trainerror,"testerror=",testerror)
   
   #analyse the results
   print(summary(fit_gam_s))
   plot.Gam(fit_gam_s, se=TRUE, col="green")
   
   return(list("dataset"=unique(trainset$name),"model"="gam_s","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_gam_s))
 }
 ```


```{r}
func_fit_rf = function(trainset,testset,factorpreds,dname) {
    formula_string = paste("count~",factorpreds,"+atemp+humidity+windspeed",sep="")
    print(formula_string)
    fit_rf = randomForest(as.formula(formula_string),data=trainset,importance=TRUE)
  print(importance(fit_rf))
  varImpPlot(fit_rf)
   trainerror = mean((trainset$count-predict(fit_rf,newdata=trainset))^2)
   testerror = mean((testset$count-predict(fit_rf,newdata=testset))^2)
  
   cat("dataset=",dname,"model=rf","trainerror=",trainerror,"testerror=",testerror)
   return(list("dataset"=unique(trainset$name),"model"="rf","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_rf))
 }
```
```{r}
func_fit_boost = function(trainset,testset,factorpreds,dname) {
    formula_string = paste("count~",factorpreds,"+atemp+humidity+windspeed",sep="")
    fit_boost = gbm(as.formula(formula_string),data=trainset,distribution="gaussian",n.trees=5000,interaction.depth=4)
print(summary(fit_boost))
 trainerror = mean((trainset$count-predict(fit_boost,newdata=trainset,n.trees=5000))^2)
 testerror = mean((testset$count-predict(fit_boost,newdata=testset,n.trees=5000))^2)
  
  cat("dataset=",dname,"model=boost","trainerror=",trainerror,"testerror=",testerror)
  
   return(list("dataset"=unique(trainset$name),"model"="boost","trainerror"=trainerror,"testerror"=testerror,"fit"=fit_boost))
 }
```

```{r 14}
 func_fit_models = function(dataset) {

  #create new variable for purposes of data partitioning
   #create stratified partition based on weather, type of day, month and hour
  dataset$partgroup = paste(dataset$weather,dataset$typeofday,dataset$mon,sep="_")
  sample =createDataPartition(dataset$partgroup,p=0.7,list=F)
  trainset = dataset[sample,]
  testset = dataset[-sample,]

  #Note: we compared model results (train and test errors) between using year and mon as numeric varaibles vs using them as factor variables (as they should be). There was a big improvement with using them as factor variables. But, for the smaller datasets where all entries all belong to the same month or year, we need to ignore the year or month factor variable.
  
  #select factor predictors for this dataset. We only want to include predictors that have at least two entries. Specifically check for season, year and mon
  dataset_name = unique(dataset$name)
   factorPreds = "weather+typeofday+wday+hour"
  #if(sum(table(dataset$season)>0)>1) {
    #factorPreds = paste(factorPreds,"season",sep="+")
  #} 
  if(sum(table(dataset$yearf)>0)>1) {
    factorPreds = paste(factorPreds,"yearf",sep="+")
  }
  if(sum(table(dataset$monf)>0)>1) {
    factorPreds = paste(factorPreds,"monf",sep="+")
  }  

  print(factorPreds)
  
  #fit lm
  #model_perf = data.frame(stringsAsFactors = FALSE)
  model_perf=list()
  model_perf = rbind(model_perf,func_fit_lm(trainset,testset,factorPreds,dataset_name))
  model_perf = rbind(model_perf,func_fit_gam_lm(trainset,testset,factorPreds,dataset_name))
  model_perf = rbind(model_perf,func_fit_gam_poly(trainset,testset,factorPreds,dataset_name))
  model_perf = rbind(model_perf,func_fit_gam_s(trainset,testset,factorPreds,dataset_name))
  model_perf = rbind(model_perf,func_fit_rf(trainset,testset,factorPreds,dataset_name))
 model_perf = rbind(model_perf,func_fit_boost(trainset,testset,factorPreds,dataset_name))
  model_perf = as.data.frame(model_perf)
  print(dim(dataset))
  print(unique(dataset$name))
  with(model_perf,print(cbind(dataset,model,trainerror,testerror)))
  #find the model that has the lowest test error
  print("model with lowest test error")
  print(model_perf[which.min(model_perf$testerror),c("dataset","model","trainerror","testerror")])
  return(model_perf[which.min(model_perf$testerror),])
 }
```


```{r 15}   

bikeTrainFitList_models = do.call(rbind,lapply(bikeTrainIncrList,FUN=function(x) func_fit_models(x)))
bikeTrainFitList_models[,1:4]
#str(bikeTrainFitList_models )
# mse_plot <- ggplot(data=bikeTrainFitList_models)+ #,aes(x=seq(nrow(bikeTrainFitList_models))) +
#    geom_line(aes(y=trainerror,colour="2")) +
#    geom_line(aes(y=testerror,colour="3")) +
#   scale_color_discrete(name = "Errors", labels = c("Train", "Test")) +
#    ggtitle("Variation in Prediction Error") +
#    xlab("Number of Months") +
#    ylab("Prediction Error: MSE") +
#   theme(legend.position="top")
#   plot(mse_plot)
plot(unlist(bikeTrainFitList_models$trainerror),type="l",col=1,xlab="Months",ylab="MSE")
lines(unlist(bikeTrainFitList_models$testerror),col=2)

```

```{r 16}
#match the model with the record
matchModelList = apply(bikeTest2, MARGIN=1, FUN=function(x) bikeTrainFitList_models[which(bikeTrainFitList_models$dataset==getYearMon(x)),])
length(matchModelList)
#str(matchModelList)
matchModelList[[1]]$dataset
testpred2 = do.call(rbind,lapply(seq(nrow(bikeTest2)),FUN=function(x) as.data.frame(c(bikeTest2$year[x],bikeTest2$mon[x],bikeTest2$mday[x],bikeTest2$hour[x], matchModelList[[x]]$dataset,bikeTest2$count[x],
      predict(matchModelList[[x]]$fit,newdata=bikeTest2[x,],n.trees=5000)),
      col.names=c("year","month","mday","hour","dataset","actual_count","predicted_count"))))

#predict with test.csv
#testpred = do.call(rbind,lapply(seq(nrow(bikeTest)),FUN=function(x) #as.data.frame(c(bikeTest$year[x],bikeTest$mon[x],
#      matchModelList[[x]]$dataset,bikeTest$count[x],
#      predict(matchModelList[[x]]$fit,newdata=bikeTest[x,],n.trees=5000)),
#      col.names=c("year","month","dataset","actual_count","predicted_count"#))))

#print(head(testpred))
print(testpred2)
str(testpred2)
testerror = mean((testpred2$actual_count-testpred2$predicted_count)^2)
testerror

```



